---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Motivation

There are 3 things I felt were missing the two wonderful packages [`XML`](https://cran.r-project.org/web/packages/XML/index.html) and [`xml2`](https://github.com/hadley/xml2):

1. A easier, more condensed way to see the structure of xml data.
  - `xml2::xml_structure` provides a way to look at the structure, but I find that it is not very easy to read and takes up too much console space
2. A quick way to determine all feasible xpaths and to identify terminal nodes. Data values of interest are contained in terminal nodes (nodes of "length zero" that do no dig any deeper). Quickly getting xpaths to the *parents* of these nodes makes extracting data much easier---and faster if you do not recursively dig deeper.
3. Other alternatives for converting xml data to data frames.
  - `XML::xmlToDataFrame` exists but it seems to always dig recursively. This leads to some crappy data frames.
  - I offer two alternatives, `xml_to_df` and `xml_dig_df`.
    - `xml_to_df` uses the `XML` and `data.table` packages
    - `xml_dig_df` is based of `xml2` and `tidyverse` packages.
    

# Installation

Run the following.

```{r, eval=FALSE}
devtools::install_github('ultinomics/xmltools')
library(xmltools)
```

# Examples

Let's set up the first example using some ebay data from the [UW XML Data Repository](http://aiweb.cs.washington.edu/research/projects/xmltk/xmldata/www/repository.html). These data come as part of the package because I dropped a really another `description` field that makes the data hard to look at. (Parses it just fine!)

```{r ex1a}
library(xmltools)

# USING ebay.xml ------------------------------------------------
# load the data
file <- system.file("extdata", "ebay.xml", package = "xmlExtras")
doc <- file %>%
  xml2::read_xml()
nodeset <- doc %>%
  xml2::xml_children() # get top level nodeset
```

## View XML trees/structures

Let's look at the structure of the data. The function

```{r ex1b, eval = F}
# `xml_view_tree` structure
## we can get a tree for each node of the doc
doc %>% 
  xml_view_tree()
doc %>% # we can also vary the depth
  xml_view_tree(depth = 2)
```

```{r ex1ba}
## easier to read and understand than `xml2::xml_structure()` and has the `depth` option
nodeset[1] %>% xml2::xml_structure()

## or, we can extract from nodesets
class(nodeset[1])
nodeset[1] %>%
  xml_view_trees()
nodeset[1] %>%
  xml_view_trees(depth=2)

## will not work with class "xml_node" (can't use lapply on those, apparently)
class(nodeset[[1]])
try(nodeset[[1]] %>%
  xml_view_tree()
)
```

## Get Terminal Nodes

Terminal nodes in XMLs are nodes that do no have any "children". These nodes contain the information we generally want to extract into a tidy data frame.

The other thing I've found myself wanting easy access to but could find no easy tool to do so was a quick way to extract all existing paths in an XML structure---especially paths to terminal nodes. The function `xml_get_paths` (or `xml_get_path`) provides this function.

```{r ex1c, eval = F}
## one can see all the paths per node of a doc
doc %>%
  xml_get_paths()
```

```{r ex1d}
## can look at one nodeset
## NOTE that nodesets can vary, so looking at one doesn't mean you'll find all feasible paths
nodeset[1] %>%
  xml_get_paths()
nodeset[1] %>%
  xml_get_paths(mark_terminal = ">>") # can mark terminal nodes

## we can find all feasible paths then collapse
terminal <- doc %>% ## get all xpaths
  xml_get_paths()
xpaths <- terminal %>% ## collapse xpaths to unique only
  unlist() %>%
  unique()
xpaths

## but what we really want is the parent node of terminal nodes.
## use the `only_terminal_parent = TRUE` to do this
terminal_parent <- doc %>% ## get all xpaths to parents of parent node
  xml_get_paths(only_terminal_parent = TRUE)
xpaths <- terminal_parent %>% ## collapse xpaths to unique only
  unlist() %>%
  unique()
xpaths

```

## Extracting XML Data to Tidy Data Frames

Next, we use the terminal xpaths above to extract the data we want.

First, I want to show the issue with using `XML::xmlToDataFrame`.

```{r ex1e}
## xmlToDataFrame works great on terminal nodes.
## we extract a data frame for each parent of terminal nodes
df0 <- lapply(xpaths, function(x) {
  doc <- file %>% XML::xmlInternalTreeParse()
  nodeset <- XML::getNodeSet(doc, x)
  XML::xmlToDataFrame(nodeset, stringsAsFactors = FALSE) %>%
    dplyr::as_data_frame()
})

# We want the data in:
#   "/root/listing/payment_types"
#   "/root/listing/shipping_info"
#   "/root/listing/buyer_protection_info"
nodeset[1] %>%
  xml_view_tree()
```

There is data contained in the terminal paths

```
/root/listing/payment_types
/root/listing/shipping_info
/root/listing/buyer_protection_info
```

with the parent node

```
/root/listing
```

But `XML::xmlToDataFrame` will keep digging into `/root/listing` and extract data from xpaths like `/root/listing/seller_info`. But it does so extracting data in a non tidy way. We can see this below comparing the data in `df0[[1]] %>% dplyr::select(seller_info)` vs `df0[[2]]`, which is data extracted from just from `/root/listing/seller_info` and deeper.

```{r ex1f}
## problem with xmlToDataFrame is it keeps digging into other nodes recursively in "/root/listing"
xpaths[1] # /root/listing is terminal parent but xmlToDataFrame keeps digging
df0[[1]] %>%
  dplyr::select(seller_info) # not good; keeps diving into other nodes but fails to separate
xpaths[2]
df0[[2]] # works because the recursive dig down hits only the terminal nodes
```


The solution that worked for me was to write a function that, by default, does not dig into non-terminal nodes. That is, when given the xpath `/root/listing`, the function will only go into terminal nodes

```
/root/listing/payment_types
/root/listing/shipping_info
/root/listing/buyer_protection_info
```

And ignore the other, non-terminal nodes unless instructed to do so otherwise by setting option `dig = TRUE`.

### `xml_to_df`

The `xml_to_df` function is built on the `XML` package and `data.table`. By default, it does not dig into non-terminal nodes (`dig = FALSE`) when given any xpath. One can pass a path to an XML file (`is_xml = FALSE`) or an already parsed XML file (`is_xml = TRUE`).

Once


```{r ex1g}
# xml_to_df (XML package based)
## does not dig by default
## use the terminal xpaths to get data frames
terminal_parent <- doc %>% ## get all xpaths to parents of parent node
  xml_get_paths(only_terminal_parent = TRUE)
terminal_xpaths <- terminal_parent %>% ## collapse terminal xpaths to unique only
  unlist() %>%
  unique()
terminal_xpaths
```

## `xml_dig_df`

The other fuction is `xml_dig_df` which is built on `xml2` and `tidyverse` packages.

The important distinction is that we first need to find all the terminal nodesets we want to parse. We find these using `xml2::xml_find_all` on each of the `terminal_xpaths` on the original parsed xml file (`doc`).

For each `terminal_nodeset`, we then apply `xml_dig_df`. For each nodeset, we will get single row of data, so we bind the results for each nodeset. The final data frame is created by column binding. (I convert all empty strings to `NA` for good measure.)

Notice that these data frames are the same. The difference is that `xml_to_df` outputs a `data.table` object and `xml_dig_df` outputs a `tibble` object.

```{r ex1h}
## we send each terminal xpath to `xml_to_df`. 
## the file source is the parsed xml object `doc`, so we set `is_xml = TRUE`
## we do no want to dig, which quickly gets us the data we want for each terminal xpath `dig = FALSE` (default)
df1 <- lapply(terminal_xpaths, xml_to_df, file = doc, is_xml = TRUE, dig = FALSE) %>%
  dplyr::bind_cols()
df1

# xml_dig_df (xml2 package based)
terminal_nodesets <- lapply(xpaths, xml2::xml_find_all, x = doc)
df2 <- terminal_nodesets %>%
  purrr::map(xml_dig_df) %>% ## does not dig by default
  purrr::map(dplyr::bind_rows) %>%
  dplyr::bind_cols() %>%
  dplyr::mutate_all(empty_as_na)
df2
```

# Example 2

Below is another example to work through. The output is hidden but copy and paste the code or find the source file in this repository `example.R`.

```{r ex2, eval = F}
# USING wsu.xml ------------------------------------------------
# larger file

# using xml_to_df
file <- "http://aiweb.cs.washington.edu/research/projects/xmltk/xmldata/data/courses/wsu.xml"
doc <- file %>%
  xml2::read_xml()
nodeset <- doc %>%
  xml2::xml_children()
length(nodeset) # lots of nodes!
nodeset[1] %>% # lets look at ONE node's tree
  xml_view_tree()

## takes a long time. likely can extract from a single node
# terminal_paths <- doc %>% ## get the xpath to parents of terminal node
#   xml_get_paths(only_terminal_parent = TRUE)

# lets assume that most nodes share the same structure
terminal_paths <- nodeset[1] %>%
  xml_get_paths(only_terminal_parent = TRUE)

xpaths <- terminal_paths %>% ## collapse xpaths to unique only
  unlist() %>%
  unique()

# xml_to_df (XML package based)
## note that we use file, not doc, hence is_xml = FALSE
df1 <- lapply(xpaths, xml_to_df, file = file, is_xml = FALSE, dig = FALSE) %>%
  dplyr::bind_cols()
df1

# xml_dig_df (xml2 package based)
## faster!
terminal_nodesets <- lapply(terminal_xpaths, xml2::xml_find_all, x = doc) # use xml docs, not nodesets! I think this is because it searches the 'root'.
df2 <- terminal_nodesets %>%
  purrr::map(xml_dig_df) %>%
  purrr::map(dplyr::bind_rows) %>%
  dplyr::bind_cols() %>%
  dplyr::mutate_all(empty_as_na)
df2

# they're the same!
identical(df1, data.table::as.data.table(df2))
```
