---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
)
```
# Motivation

There are 3 things I felt were missing the two wonderful packages [`XML`](https://cran.r-project.org/web/packages/XML/index.html) and [`xml2`](https://github.com/hadley/xml2):

1. A easier, more condensed way to see the structure of xml data.
  - `xml2::xml_structure` provides a way to look at the structure, but I find that it is not very easy to read and takes up too much console space
2. A quick way to determine all feasible xpaths and to identify terminal nodes. Data values of interest are contained in terminal nodes (nodes of "length zero" that do no dig any deeper). Quickly getting xpaths to the *parents* of these nodes makes extracting data much easier---and faster if you do not recursively dig deeper.
3. Other alternatives for converting xml data to data frames.
  - `XML::xmlToDataFrame` exists but it seems to always dig recursively. This leads to some crappy data frames.
  - I offer two alternatives, `xml_to_df` and `xml_dig_df`.
    - `xml_to_df` uses the `XML` and `data.table` packages
    - `xml_dig_df` is based of `xml2` and `tidyverse` packages.
    

# Installation

```{r}
devtools::install_github('ultinomics/xmlExtras')
library(xmlExtras)

# USING ebay.xml ------------------------------------------------
# learn about the structure
file <- 'data-raw/ebay.xml'
doc <- file %>%
  xml2::read_xml()
tree <- doc %>%
  xml_get_trees()
nodeset <- doc %>%
  xml2::xml_children()

## structure
tree[[1]] %>% xml_view_tree()
nodeset[[1]] %>% xml2::xml_structure()
terminal <- doc %>% ## get the xpath to parents of terminal node
  xml_get_paths(only_terminal_parent = TRUE)
xpaths <- terminal %>% ## collapse xpaths to unique only
  unlist() %>%
  unique()

# xmlToDataFrame works great on terminal nodes.
# problem with xmlToDataFrame is it keeps digging.
df0 <- lapply(xpaths, function(x) {
  doc <- file %>% XML::xmlInternalTreeParse()
  nodeset <- XML::getNodeSet(doc, x)
  XML::xmlToDataFrame(nodeset, stringsAsFactors = FALSE) %>%
    dplyr::as_data_frame()
})

doc %>%
  xml_get_paths(mark_terminal = ">>") %>%
  '[['(1)

# want just:
#   "/root/listing/payment_types"
#   "/root/listing/shipping_info"
#   "/root/listing/buyer_protection_info"
xpaths[1] # /root/listing is terminal parent but xmlToDataFrame keeps digging
df0[[1]] # not good; keeps diving into other nodes
xpaths[2]
df0[[2]] # good because it could dive further

# xml_to_df (XML package based)
## does not dig by default
## use the terminal xpaths to get data frames
df1 <- lapply(xpaths, xml_to_df, file = doc, is_xml = TRUE, dig = FALSE) %>%
  dplyr::bind_cols()
df1

# xml_dig_df (xml2 package based)
## does not dig by default
tree[[1]] %>% xml_view_tree()
terminal_nodesets <- lapply(xpaths, xml2::xml_find_all, x = doc)
df2 <- terminal_nodesets %>%
  purrr::map(xml_dig_df) %>%
  purrr::map(dplyr::bind_rows) %>%
  dplyr::bind_cols()


# USING wsu.xml ------------------------------------------------
# larger file

# using xml_to_df
file <- 'data-raw/wsu.xml'
doc <- file %>%
  xml2::read_xml()
nodeset <- doc %>%
  xml2::xml_children()
length(nodeset) # lots of nodes!
nodeset[1] %>% # lets look at ONE node's tree
  xml_view_tree()

## takes a long time. likely can extract from a single node
# terminal_paths <- doc %>% ## get the xpath to parents of terminal node
#   xml_get_paths(only_terminal_parent = TRUE)

# lets assume that most nodes share the same structure
terminal_paths <- nodeset[1] %>%
  xml_get_paths(only_terminal_parent = TRUE)

xpaths <- terminal_paths %>% ## collapse xpaths to unique only
  unlist() %>%
  unique()

# xml_to_df (XML package based)
## note that we use file, not doc, hence is_xml = FALSE
df1 <- lapply(xpaths, xml_to_df, file = file, is_xml = FALSE, dig = FALSE) %>%
  dplyr::bind_cols()
df1

# xml_dig_df (xml2 package based)
## faster!
terminal_nodesets <- lapply(xpaths, xml2::xml_find_all, x = doc) # use xml docs, not nodesets! I think this is because it searches the 'root'.
df2 <- terminal_nodesets %>%
  purrr::map(xml_dig_df) %>%
  purrr::map(dplyr::bind_rows) %>%
  dplyr::bind_cols() %>%
  dplyr::mutate_all(empty_as_na) # kill of NAs
df2

```
  
